---
title: "Fitting SARS-CoV-2 variant frequency growth using (multinomial) logistic regression"
author: "Marlin Figgins"
output: html_document
---

```{r}
library(tidyverse)
```


## Loading and processing data

```{r}
raw_df = read.csv("../metadata-usa.clean.csv") %>% select(date, pango_lineage, division)
```

```{r}
variants.list <- c("B.1.1.7", "P.1", "B.1.351", "B.1.526", "B.1.617") 

clean_df <- raw_df %>%
  mutate(date = as.Date(date)) %>% # Use datetime
  mutate(pango_lineage = case_when(pango_lineage %in% variants.list ~ pango_lineage, # pango_lineage to variants we care about
                                   TRUE ~ "other")) %>%
  count(date, pango_lineage, division) %>% # Count occurrence of each lineage and division
  group_by(date, division) %>%
  mutate(sequence_total = sum(n)) %>% # Compute total sequences observed on that day in that place
  ungroup() %>%
  mutate(freq = n / sequence_total) # Compute total frequency of lineage for each day and division

clean_df_wide <- clean_df %>% # Pivot to wide format for passing to models
  rename(sequences = n) %>%
  pivot_wider(names_from = pango_lineage, values_from = c(sequences, freq), values_fill = 0) 
```

```{r}
clean_df_wide
```

# Preliminary plots

```{r}
## Quick look at sequences 
clean_df  %>%
  ggplot(aes(x=date, y = n, id = division)) +
  geom_line() +
  facet_wrap(~pango_lineage)
```

```{r}
# Look at lineage frequencies in a few states
clean_df %>%
  filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x=date, y = freq, color = pango_lineage)) +
  geom_line() +
  facet_wrap(~division)
```


## Fitting linear model on the logit scale to frequncies

We'll model each frequency trajectory over time as a logistic function, so that the frequency of variant $v$ on day $t$ is given by

$$
f_{v}(t) = \frac{\exp(\beta_{0,v} + \beta_{1,v}t)}{1 + \exp(\beta_{0,v} + \beta_{1,v}t)}.
$$
We implement this model as a linear regression on the logit of the frequency, so that our observed frequencies $y_{v,t}$ are 

$$
\text{logit} (y_{v,t}) = \beta_{0,v} + \beta_{1,v} t  + \epsilon, \quad \epsilon \sim \text{Normal}(0, \sigma) .
$$

```{r}
library(broom)
library(modelr)

test_results <- clean_df %>%
  mutate(logit_freq = qlogis(freq)) %>% # Get frequency on logit scale
  filter(is.finite(logit_freq)) %>% # Filter out extreme values
  nest(-division, -pango_lineage) %>% # Fit linear model by division and lineage on logit scale
  mutate(model = map(data, ~lm(logit_freq ~ time(date), data = .)), 
         data = map2(.x = data, .y = model, ~add_predictions(.x, .y, var = "logit_pred"))) %>% # Get predictions
  unnest(data) %>%
  mutate(pred = plogis(logit_pred)) # Transform logit predictions to [0,1]

test_results
```

```{r}
#clean_df %>%
#  mutate(logit_freq = qlogis(freq)) %>% # Get frequency on logit scale
#  filter(is.finite(logit_freq)) %>% # Filter out extreme values
#  nest(-division, -pango_lineage) %>% # Fit linear model by division and lineage on logit scale
#  mutate(model = map(data, ~lm(logit_freq ~ time(date), data = .)), 
#         data = map2( .x = model, .y = data, ~augment(.x, data = .y))) %>% # Get predictions
 # unnest(data) %>%
#  mutate(pred = plogis(.fitted)) # Transform logit predictions to [0,1]
```
## Plotting results

```{r, fig.height=10, fig.width=14}
# Plot fits in ggplot1
test_results %>%
  filter(division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot( aes(x = date, color = pango_lineage)) +
  geom_point(aes(y = freq, size = sequence_total)) +
  scale_size_continuous(range = c(1, 3)) +
  geom_line(aes(y = pred), lwd = 1.5) +
  facet_wrap(~division) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```
```{r, fig.height=5, fig.width=7}
# Plot fits in ggplot1
test_results %>%
  filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot( aes(x = date, color =division )) +
  geom_point(aes(y = freq, size = sequence_total)) +
  scale_size_continuous(range = c(1, 3)) +
  geom_line(aes(y = pred), lwd = 1.5) +
  facet_wrap(~pango_lineage) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```

## Fitting multinomial logistic

We now fit a multinomial regression to all $N$ lineages of interest within a division simultaneously. This is assuming that frequencies of each lineage in the population evolve as

$$
f_{v}(t) = \frac{\exp(\beta_{0,v} + \beta_{1, v} t)}{\sum_{i=1}^N \exp(\beta_{0, i} + \beta_{1, i} t)}
$$
We can re-frame this model as a model on the actual case growth of infections with specific variants as

$$
f_{v}(t) = \frac{p_{v}\exp(r_v t)}{\sum_{i=1}^{N} p_i \exp(r_{i}t)} = \frac{I_{v}(t)}{\sum_{i=1}^{N} I_i(t)},
$$
so that if $I_{v}(t)$ is the number of individuals infected with variant $v$, then $p_v = \exp(\beta_{0,v})$ is the number infected at time $0$ and $r_{v} = \beta_{1,v}$ is the exponential growth rate of infections with variant $v$. In R, we can implement this model using `nnet`.

```{r}
library(lubridate)
library(nnet)
```

```{r}

get_division_mat <- function(tbl){
  return(tbl %>%
  select(starts_with("sequences")) %>% # Get sequence counts
  rename_all(~str_replace(.,"^sequences","pred_freq")) %>% # Rename variable for later
  as.matrix())
}

get_division_times <- function(tbl){
  dt_mat <- tbl %>%
  select(date) %>%
  mutate(times = yday(date),
         times = (times - min(times)) / (max(times) - min(times))) # Transform times for easy fitting
  return(dt_mat$times)
}

multinomial_frame <- clean_df_wide %>%
  nest(-division) %>% # Nest by division
  mutate(division_mat = map(data, ~get_division_mat(.)), # Get counts 
         division_times = map(data, ~get_division_times(.)), # Get times
         model = map2(.x = division_mat, .y = division_times, ~multinom(.x ~ .y, trace = FALSE)),
         predictions = map(model, ~ predict(., type = "prob")),  # Get predictions
         data = map2(.x = data, .y = predictions, ~ cbind(.x, .y))) %>%  # Augment original data with predictions
  select(division, data, model) %>%
  unnest(data)

multinomial_frame
```


```{r}
# Reduce data to long format
sequence_n <- multinomial_frame %>%
  select(division, date, sequence_total, starts_with("sequences")) %>%
  pivot_longer(starts_with("sequences"), names_to = "pango_lineage", values_to = "sequence_n") %>%
  mutate(pango_lineage = str_replace(pango_lineage,"^sequences_",""))

obs_freqs <- multinomial_frame %>%
  select(division, date, sequence_total, starts_with("freq")) %>%
  pivot_longer(starts_with("freq"), names_to = "pango_lineage", values_to = "obs_freq") %>%
  mutate(pango_lineage = str_replace(pango_lineage,"^freq_",""))

pred_freq <- multinomial_frame %>%
  select(division, date, sequence_total, starts_with("pred_freq")) %>%
  pivot_longer(starts_with("pred_freq"), names_to = "pango_lineage", values_to = "pred_freq") %>%
  mutate(pango_lineage = str_replace(pango_lineage,"^pred_freq_",""))

multinomial_longer <- sequence_n %>% left_join(obs_freqs) %>% left_join(pred_freq)

multinomial_longer
```

```{r}
# Plot samples in ggplot2
multinomial_longer %>%
  filter(division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x = date, y= obs_freq, fill = pango_lineage)) +
  geom_area() +
  facet_wrap(~division) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```


```{r}
# Plot fits in ggplot2

# Can add error bars to the data using Jeffrey's interval later
multinomial_longer %>%
  filter(division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x = date, y= pred_freq, fill = pango_lineage)) +
  geom_area() +
  facet_wrap(~division) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```


```{r, fig.height=10, fig.width=14}
# Plot fits in ggplot2

# Can add error bars to the data using Jeffrey's interval later
states_plt <- multinomial_longer %>%
  filter(division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x = date, color = pango_lineage)) +
  geom_point(aes(y = obs_freq, size = sequence_total)) +
  scale_size_continuous(range = c(1, 3)) +
  geom_line(aes(y = pred_freq), lwd = 1.5) +
  facet_wrap(~division) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())

#ggsave("multinomial-test.png", plot = states_plt)
states_plt
```

```{r}
multinomial_longer %>%
  filter(division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x = date, color = division)) +
  geom_point(aes(y = obs_freq, size = sequence_total)) +
  scale_size_continuous(range = c(1, 1.3)) +
  geom_line(aes(y = pred_freq), lwd = 1.5) +
  facet_wrap(~pango_lineage) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```

Question: Are there any notable trends in $r_v / r_{\text{other}}$ across different division? 
- This will be manageable by looking at the model objects to get their parameters, comparing those of baseline category to all others. 

Question: Is this model usable for prediction? What happens if we fit all but a months data, then try to forecast that month?
- Divide data into training and test set. Evaluate performance of multinomial model trained on training data on the testing data.

Question: Where might the exponential growth assumption become an issue?


## Loading CDC data

```{r}
cdc_data <- read.csv("https://data.cdc.gov/api/views/9mfq-cb36/rows.csv?accessType=DOWNLOAD") %>% 
  select(submission_date, state, new_case) %>% # Get columns of interest
  rename(date = submission_date, cases = new_case, division = state) %>% # Rename to match GISAID
  mutate(date = as.Date(date, "%m/%d/%Y"), # Change date and division formatting to match
         division = state.name[match(division, state.abb)]) %>%
  filter(!is.na(division))# Filter for divisions which weren't matched
```

```{r}
full_df <- multinomial_longer %>% left_join(cdc_data) %>%
  mutate(pred_cases = pred_freq * cases)
```


```{r,  fig.height=10, fig.width=14}
full_df %>%
  filter(division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "Illinois", "Washington")) %>%
  ggplot(aes(x = date, fill = pango_lineage)) +
  geom_col(aes(y = pred_cases)) +
  facet_wrap(~division) +
  #coord_cartesian(ylim = c(0, 10000)) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```

## Export .csv for use else where

```{r}
write.csv(full_df, "usa-multinomial-fit.csv")
```