---
title: "Untitled"
author: "Marlin Figgins"
date: "6/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(EpiEstim)
library(tidyverse)


```

## Example using EpiEstim with sample data

```{r}
# Code sample is taken from https://cran.r-project.org/web/packages/EpiEstim/vignettes/demo.html

## load example data
data(Flu2009)

# Estimate R using Flu2009 incidence data.
res_parametric_si <- estimate_R(Flu2009$incidence, 
                                method="parametric_si", # Serial interval is described by Gamma distribution
                                config = make_config(list( # The config variable contains the SI mean and standard deviation
                                  mean_si = 2.6, 
                                  std_si = 1.5))
)

res_parametric_si$R # the model object contains a data frame with info on R
```

```{r}

process_rt <- function(x){
  t_start <- x$R$t_start
  t_end <- x$R$t_end
  dates <- x$dates
  mean_posterior <- x$R[, "Mean(R)"]
  quantile_0.025_posterior <- x$R[, "Quantile.0.025(R)"]
  quantile_0.975_posterior <- x$R[, "Quantile.0.975(R)"]
  
  return(data.frame(
   # start = dates[t_start], 
    dates = dates[t_end], 
    meanR = mean_posterior,
    lower = quantile_0.025_posterior,
    upper = quantile_0.975_posterior))
}
  
test_rt <- process_rt(res_parametric_si)

ggplot(test_rt, aes(dates, meanR)) +
          geom_ribbon(aes(ymin = lower, ymax = upper, fill = "95%CrI")) +
          geom_line(color = "black") +
          geom_hline(yintercept = 1, linetype = "dotted") +
          ggtitle("Estimated R")

```

EpiEstim by default infers R in terms of sliding windows. We can average the R results across windows containing each time point as follows:

```{r}
# Processing model R
res_parametric_si$R %>%
  mutate(count = t_end - t_start + 1) %>% # Add row for each individual t
  uncount(count, .id = "t") %>% 
  mutate(t = t_start + t - 1) %>% # Shift to correspond with dates
  group_by(t) %>%  # Group by each t
  summarize(rt_mean = mean(`Mean(R)`), # Take mean across all rts for that date
            rt_median = mean(`Median(R)`), 
            rt_quantile.25 = mean(`Quantile.0.25(R)`),
            rt_quantile.75 = mean(`Quantile.0.75(R)`)) %>%
  add_row(t = 1, rt_mean = 1, rt_median = 1, rt_quantile.25 = 1, rt_quantile.75  = 1) %>% # Add row for first t and Rt = 1
  ggplot(aes(x = t)) +
  geom_ribbon(aes(ymin = rt_quantile.25, ymax = rt_quantile.75), fill = "grey70") +
  geom_line(aes(y = rt_median))
```


```{r}
# Importing previous output
Rt_input_df <- read.csv("usa-multinomial-fit.csv") %>%
  select(date, division, pango_lineage, pred_cases) %>% # Selecting dates, division, lineage, and predicted_cases of that linege
  filter(!is.na(pred_cases)) %>%
  mutate(date = as.Date(date)) %>%
  rename(I = pred_cases, dates = date) %>% # Rename columns for EpiEstim
  #mutate_at(vars(I), ~ as.integer(round(.x))) %>% # EpiEstim requires integer incidence
  mutate(I = abs(I)) # Make sure everything is positive
```

```{r}
# Replace these with estimates from a paper of some kind
config <- make_config(list(mean_si = 5.2, std_si = 2.8)) # using Gayani estimate

get_Rt <- function(incidence_df){
    res <- incidence_df %>%
      arrange(dates) %>%
      estimate_R(method="parametric_si", config = config)
    return(res)
}

clean_R_values <- function(Rt_model){
   Rt_model$R  %>%
    mutate(count = t_end - t_start + 1) %>% # Add row for each individual t
    uncount(count, .id = "t") %>% 
    mutate(t = t_start + t - 1) %>%
    group_by(t) %>%  # Group by each t
    summarize(rt_mean = mean(`Mean(R)`), # Take mean across all Rts for that date
              rt_median = mean(`Median(R)`), 
              lower = mean(`Quantile.0.025(R)`),
              upper = mean(`Quantile.0.975(R)`)) %>%
    add_row(t = 1, rt_mean = 1, rt_median = 1, rt_quantile.25 = 1, rt_quantile.75 = 1) %>% # Add row for first t and Rt = 1
    arrange(t) %>% # Make sure t's are ordered
     add_column(dates = Rt_model$dates)
}

process_rt <- function(x){
  t_start <- x$R$t_start
  t_end <- x$R$t_end
  dates <- x$dates
  mean_posterior <- x$R[, "Mean(R)"]
  median_posterior <- x$R[, "Mean(R)"]
  quantile_0.025_posterior <- x$R[, "Quantile.0.025(R)"]
  quantile_0.975_posterior <- x$R[, "Quantile.0.975(R)"]
  
  return(data.frame(
   # start = dates[t_start], 
    dates = dates[t_end], 
    rt_mean = mean_posterior,
    rt_median = median_posterior,
    lower = quantile_0.025_posterior,
    upper = quantile_0.975_posterior))
}

augment_original_data <- function(data, Rt_model){
  data %>% left_join(process_rt(Rt_model))
}

division_of_interest <- c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")
Rt_output_df <- Rt_input_df %>%
  filter(division %in% division_of_interest) %>%
  nest(-division, -pango_lineage) %>% 
  mutate(model = map(data, ~get_Rt(.)),  # Compute Rt
         aug_data = map2(.x = data, .y = model, ~ augment_original_data(.x, .y))) %>% # Process Rt data
  select(division, pango_lineage, aug_data) %>%
  unnest(aug_data)
```


```{r, fig.height=10, fig.width=10}
Rt_output_df %>%
  ggplot(aes(x = dates,id = division)) +
  geom_hline(yintercept = 1,  linetype="dashed") +
  geom_line(aes(y = rt_median, color = pango_lineage),) + 
  facet_wrap( ~ division) + 
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = pango_lineage), alpha = 0.2) +
  ylim(NA, 2)
```


Question: Does assuming that we have a multinomial regression fit on the frequencies enforce a condition on R being just a constant multiple of all other lineages? Or some other constraint?

That is, is there an easy relationship between relative Rt and the variant specific exponential growth rates $r_{v}$? Olympics paper should be a good starting point for this.

```{r, fig.height=10, fig.width=10}
Rt_output_df %>%
  ggplot(aes(x = dates, id = pango_lineage)) +
  geom_hline(yintercept = 1,  linetype="dashed") +
  geom_line(aes(y = rt_median, color = division)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = division), alpha = 0.1) +
  facet_wrap( ~ pango_lineage) + 
  ylim(NA, 2)
```