---
title: "06-21-21-COV-logisticReg.Rmd"
author: "Eslam Abousamra"
date: "6/22/2021"
output: word_document
---
---
title: "state-estimate-nnetpackage"
author: "Eslam Abousamra"
date: "6/21/2021"
output: word_document
---


# growthrate estimate using nnet package

# Installing packages

```{r}
library(nnet)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)

```


# Reading and tidying the dataset

```{r}

covmeta_raw = read.csv("metadata-usa.csv")


#Create new dataframe with date, state, and lineage
init_df= data.frame(as.Date(covmeta_raw$date,"%m/%d/%y"), covmeta_raw$division, covmeta_raw$pango_lineage)

names(init_df)

init_df = init_df %>% 
  rename(
    date = as.Date.covmeta_raw.date....m..d..y..
  )

#Count frequency of different lineages in different states every day
mod_df = init_df %>%
  count(date, covmeta_raw.pango_lineage,covmeta_raw.division)%>%
  group_by(date, covmeta_raw.division) %>%
  mutate(totalcount = sum(n))%>% 
  ungroup() %>% 
  mutate(freq = n/totalcount)
?count
mod_df


mod_df_wide <- mod_df %>% # Pivot to wide format for passing to models
  rename(sequences = n) %>%
  pivot_wider(names_from = covmeta_raw.pango_lineage, values_from = c(sequences, freq), values_fill = 0) 





```


# Initial Visualization


```{r}
#Filter states and Lineages of Interest
mod_df %>%
  filter(covmeta_raw.division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%

   filter(covmeta_raw.pango_lineage %in% c("B.1.1.7", "P.1","B.1.526")) %>%
  
 ggplot(aes(x=time(date), y = freq, color = covmeta_raw.pango_lineage)) + 
  geom_line(group=1) +
  facet_wrap(~covmeta_raw.division)
```


# Fitting the logit regression


```{r}
library(modelr)

#logit transformation

mod_df = mod_df %>% mutate(tlogit = qlogis(freq))


mod_df2 = mod_df %>% 
  
  filter(is.finite(tlogit)) %>% #Removing inf values
  
  nest(-covmeta_raw.division, -covmeta_raw.pango_lineage) %>% 
  
  mutate(fit1 = map(data, ~lm(tlogit ~ time(date), data = .)), data = map2(.x = data, .y = fit1, ~add_predictions(.x, .y, var = "logit_pred"))) %>%
  
   unnest(data) %>%
  mutate(pred = plogis(logit_pred))
      
mod_df2

  







```



# Visualizing the predicted fit 


```{r}
mod_df2 %>%
  filter(covmeta_raw.division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%  filter(covmeta_raw.pango_lineage %in% c("B.1.1.7", "P.1","B.1.526")) %>%
  ggplot( aes(x = date, color = covmeta_raw.pango_lineage)) +
  geom_point(aes(y = freq, size = totalcount)) +
  scale_size_continuous(range = c(1, 3)) +
  geom_line(aes(y = pred), lwd = 1.5) +
  facet_wrap(~covmeta_raw.division) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```


#Multinomial Logistic fit (nnet Package)

```{r}
library(lubridate)
library(nnet)

#Create matrix tibble for states 

get_division_mat <- function(tbl){
  return(tbl %>%
  select(starts_with("sequences"), -totalcount) %>% # Get sequence counts
  rename_all(~str_replace(.,"^sequences","pred_freq")) %>% # Rename variable for later
  as.matrix())
}


get_division_times <- function(tbl){
  dt_mat <- tbl %>%
  select(date) %>%
  mutate(times = yday(date),
         times = (times - min(times)) / (max(times) - min(times))) # Transform times for easy fitting
  return(dt_mat$times)
}

class(mod_df$covmeta_raw.division)

#Fitting the multinomial frame

multinomial_frame <- mod_df_wide %>%
  nest(-covmeta_raw.division) %>%

# Nest by division
  mutate(division_mat = map(data, ~get_division_mat(.)), # Get counts 
         division_times = map(data, ~get_division_times(.)), # Get times
         model = map2(.x = division_mat, .y = division_times, ~multinom(.x ~ .y, trace = FALSE, MaxNWts = 100000)),predictions = map(model, ~ predict(., type = "prob")),  # Get predictions
         data = map2(.x = data, .y = predictions, ~ cbind(.x, .y))) %>% 
  
  # Augment original data with predictions
  select(covmeta_raw.division, data, model) %>%
  unnest(data)

multinomial_frame
```

```{r}
# Reduce data to long format
sequence_n <- multinomial_frame %>%
  select(covmeta_raw.division, date, totalcount, starts_with("sequences")) %>%
  pivot_longer(starts_with("sequences"), names_to = "pango_lineage", values_to = "sequence_n") %>%
  mutate(pango_lineage = str_replace(pango_lineage,"^sequences_",""))
obs_freqs <- multinomial_frame %>%
  select(covmeta_raw.division, date, totalcount, starts_with("freq")) %>%
  pivot_longer(starts_with("freq"), names_to = "pango_lineage", values_to = "obs_freq") %>%
  mutate(pango_lineage = str_replace(pango_lineage,"^freq_",""))
pred_freq <- multinomial_frame %>%
  select(covmeta_raw.division, date, totalcount, starts_with("pred_freq")) %>%
  pivot_longer(starts_with("pred_freq"), names_to = "pango_lineage", values_to = "pred_freq") %>%
  mutate(pango_lineage = str_replace(pango_lineage,"^pred_freq_",""))
multinomial_longer <- sequence_n %>% left_join(obs_freqs) %>% left_join(pred_freq)

multinomial_longer












```


# Plotting the multinomial Regression by state (division)

```{r}


states_plt <- multinomial_longer %>%
  filter(covmeta_raw.division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>% filter(pango_lineage %in% c("B.1.1.7", "P.1","B.1.526")) %>%
  ggplot(aes(x = date, color = pango_lineage)) +
  geom_point(aes(y = obs_freq, size = totalcount)) +
  scale_size_continuous(range = c(1, 3)) +
  geom_line(aes(y = pred_freq), lwd = 1.5) +
  facet_wrap(~covmeta_raw.division) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())

#ggsave("multinomial-test.png", plot = states_plt)
states_plt




```


# Plotting the multinomial Regression by Lineage



```{r}
multinomial_longer %>%
  filter(covmeta_raw.division %in% c("California", "Tennessee", "Oregon", "Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  filter(pango_lineage %in% c("B.1.1.7", "P.1","B.1.526")) %>%
  ggplot(aes(x = date, color = covmeta_raw.division)) +
  geom_point(aes(y = obs_freq, size = totalcount)) +
  scale_size_continuous(range = c(1, 1.3)) +
  geom_line(aes(y = pred_freq), lwd = 1.5) +
  facet_wrap(~pango_lineage) +
  theme(legend.title = element_blank(), 
        axis.title.y=element_blank(),
        axis.title.x=element_blank())
```

