---
title: "06-28-21-Rt_EpiEstim_1"
author: "Eslam Abousamra"
date: "6/28/2021"
output: word_document
---


### @article{Cori2013, author={Cori, A and Ferguson, NM and Fraser, C and Cauchemez, S},
###year={2013},
###title={{A New Framework and Software to Estimate Time-Varying Reproduction Numbers During ###Epidemics}},
###journal={Am. J. Epidemiol.},
###doi={10.1093/aje/kwt133},
###}





```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(EpiEstim)
library(cowplot)
library(altair)
library(vegalite)

```


# Importing and reading tidy dataset
```{r}
full_data = read.csv("Rt_cdc&GISAID_Jul_21.csv") %>%
  select(-X) %>% 
  rename(pango_lineage = Pango.lineage) 

```


# Initial exploratory analysis 
```{r}

#Exploratory analysis (predicted frequencies)

#Visualizing frequencies for lineages of concern
plot1 = full_data %>%
  filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x= as.Date(date),y = pred_freq,  fill = division)) + 
  geom_area() + 
  ylab("Predicted frequencies") +
  theme_classic() +
  facet_wrap(~pango_lineage) +
  theme(legend.title = element_blank(),
        axis.title.x=element_blank(),
        strip.background =element_blank())
  
plot1

plot1_2 = full_data %>%
  #filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x= as.Date(date),y = pred_freq,  fill = pango_lineage)) + 
  geom_area() + 
  ylab("Predicted frequencies") +
  theme_classic() +
  facet_wrap(~division) +
  theme(legend.title = element_blank(),
        axis.title.x=element_blank(),
        strip.background =element_blank())
plot1_2


plot1_4 = full_data %>%
  filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) %>%
  ggplot(aes(x= as.Date(date),y = pred_freq,  color = pango_lineage)) + 
  geom_line() + 
  ylab("Predicted frequencies") +
  theme_classic() +
  facet_wrap(~division) +
  theme(legend.title = element_blank(),
        axis.title.x=element_blank(),
        strip.background =element_blank())
plot1_4



#Visualizing predicted cases for lineages of concern
plot2 = full_data %>%
  filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington"))  %>%
  ggplot(aes(x=as.Date(date), y = pred_cases, color = division)) + 
  #geom_point(size = 0.5, alpha = 1) +
  geom_area(aes(fill = division), alpha = 0.5) +
  ylab("Case count") +
  theme_classic() +
  facet_wrap(~pango_lineage) +
  theme(legend.title = element_blank(),
        axis.title.x=element_blank(),
        strip.background =element_blank()) +
  ylim(0, 18000)

plot2


plot2 = full_data %>%
  filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington"))  %>%
  ggplot(aes(x=as.Date(date), y = pred_cases, color = division)) + 
  #geom_point(size = 0.5, alpha = 1) +
  geom_area(aes(fill = division), alpha = 0.5) +
  ylab("Case count") +
  theme_classic() +
  facet_wrap(~pango_lineage) +
  theme(legend.title = element_blank(),
        axis.title.x=element_blank(),
        strip.background =element_blank()) +
  ylim(0, 18000)











```


# Estimating Rt using EpiEstim Package

## Initial Analysis of Rt 


# Estimating Rt for nested group of states and lineages

## Preparing dataset and creating helper functions to process

```{r}



#Customizing variables to pass in the EpiEstim Package
Rt_data = full_data %>% 
  select(date, pango_lineage, division, pred_cases) %>% 
  rename(I = pred_cases) %>% 
  rename(dates = date) %>% 
  mutate(dates = as.Date(dates)) %>%
  mutate(I = abs(I)) #estimate absolute value of Incidence 



#Create Rt estimate function
#Serial interval distribution (Gamma) parameters obtained from literature

get_Rt <- function(tbl){ #Argument pass tibble (data)
  Rt = estimate_R(tbl,
             method = "parametric_si", #Specifying method of serial interval estimation
           config = make_config(list(
             mean_si = 5.2,
             std_si = 2.8)))
  return(Rt)
}


# Processing Rt and extracting mean R value

process_rt = function(x){
  t_start = x$R$t_start
  t_end = x$R$t_end
  dates = x$dates
  mean_posterior = x$R[, "Mean(R)"]
  quantile_0.025_posterior <- x$R[, "Quantile.0.025(R)"]
  quantile_0.975_posterior <- x$R[, "Quantile.0.975(R)"]
  median_posterior = x$R[, "Median(R)"]
  return(data.frame(
    dates = dates[t_end],
    meanR = mean_posterior,
    medianR =median_posterior,
    lower = quantile_0.025_posterior,
    upper = quantile_0.975_posterior))
}


#nest division, lineage to pass in Rt estimate function

Rt_output = Rt_data %>% 
  nest(-division, -pango_lineage) %>%
  mutate(Rt_est = map(data, ~get_Rt(.)) , 
         R_value = map(Rt_est, ~process_rt(.)),
         R_mod = map2(.x = data, .y= R_value, ~left_join(.x,.y))) %>% 
  select(division, pango_lineage, R_mod) %>%
  unnest(R_mod)

write.csv(Rt_output, "Rt_EpiEstim_est.csv")


```



# Visualization
```{r}
#Plotting Rt
filter(division %in% c("Michigan", "New York", "Texas", "North Carolina", "Illinois", "Washington")) 

Rt_output %>%
  filter(division %in% c("Washington", "New York", "Texas")) %>%
ggplot( aes(x = dates, y = meanR)) +
  ylab("Mean Rt") +
  facet_wrap(vars(division)) +
  geom_line(aes(colour =pango_lineage)) +
  #geom_smooth(se = FALSE) + 
  geom_hline(yintercept=1, linetype=3, col = 'darkslategrey', size= 0.7) +
  theme_classic() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = pango_lineage), alpha = 0.3)+ 
  theme(legend.title = element_blank(),
        axis.title.x=element_blank(),
        strip.background =element_blank(),
        aspect.ratio = 0.65)+
  ylim(NA,10)
Rt_output 


head(Rt_output)
Rt_output %>%
 mutate(upper = max(1:2, upper))

#Renaming lineages to greek letters
Rt_output$pango_lineage = gsub("B.1.1.7","Alpha",Rt_output$pango_lineage)
Rt_output$pango_lineage = gsub("P.1","Gamma",Rt_output$pango_lineage)
Rt_output$pango_lineage = gsub("B.1.351","Beta",Rt_output$pango_lineage)
Rt_output$pango_lineage = gsub("B.1.526","Iota",Rt_output$pango_lineage)
Rt_output$pango_lineage = gsub("B.1.617.2","Delta",Rt_output$pango_lineage)



head(Rt_output)

#vof = c("B.1.1.7", "P.1", "B.1.351", "B.1.526", "B.1.617.2") 
epie_rt = Rt_output %>%
  filter(division %in%c("Washington", "New York", "Texas"))  %>%
  filter(upper <2) %>%
ggplot(aes(x = dates, y = medianR)) +
  ylab("Median Rt") +
  facet_wrap(vars(division)) +

  geom_line(aes(color =pango_lineage), size = 1.5) +
  scale_fill_manual(values=c("#4F34D2","#5299e0","#b5d765","#ff5a30","#83cd93",
"#dadfe2"))  +
       scale_color_manual(values=c("#4F34D2","#5299e0","#b5d765","#ff5a30","#83cd93",
"#dadfe2"))  +
  #geom_smooth(se = FALSE) + 
  geom_hline(yintercept=1, linetype=4, col = 'darkslategrey', size= 1) +
  theme_classic() +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = pango_lineage), alpha = 0.17)+ 
  theme(legend.title = element_blank(),
        legend.position = "none",
        axis.title.x=element_blank(),
        strip.background =element_blank(),
        strip.text = element_text(size=15)) +
coord_cartesian(ylim = c(NA,2), clip = "on")

epie_rt



ggsave("../epiestim_r.png", plot = epie_rt,width = 3000,
  height = 1200, units= ("px"))



```


#Arranged plots

```{r}

ggarrange(counts_plot, NULL, pred_freq, Observed_freqs , 
          ncol = 2, nrow = 2,  align = "hv", 
          widths = c(2, 1), heights = c(1, 2),
          common.legend = TRUE)





```


