---
title: "5/18/2021-Initial-Frequency-estimate-B.1.1.7-State"
author: "Eslam Abousamra"
date: "5/18/2021"
output:
  word_document: default
  pdf_document: default
---


```{r}

library(tidyverse)
library(lubridate)
library(dplyr)
library(zoo)
library(TTR)

```

# Reading Dataset
## COVID-19 Lineages across the US
### Purpose of the analysis is to visualize the frequency of B.1.1.7 Lineage in the US


```{r}
covmeta_raw = read.csv("metadata-usa.csv")
View(covmeta_raw)
covtry= data.frame(covmeta_raw$date, covmeta_raw$division, covmeta_raw$pango_lineage)
head(covtry)

```

# Frequency of B.1.1.7 Lineage across US

### Estimation of the ratio of occurences of B.1.1.7 lineage across the US
### and plotting the smoothed 7-day moving average over time


```{r}
df = covtry 
strain1= table(df$covmeta_raw.pango_lineage=="B.1.1.7")
new_df <- df %>%
  drop_na() %>% # if we want "ignore" NAs when counting
  group_by(covmeta_raw.date) %>%
  summarise(ratio = sum(covmeta_raw.pango_lineage=="B.1.1.7")/(cumsum(n())+0.0001))
ggplot(new_df, aes(x=covmeta_raw.date, y=ratio*100)) + geom_point() + labs(x = "Date",y ="Lineage")

new_df %>% group_by(covmeta_raw.date)
sorted_data = new_df[order(as.Date(new_df$covmeta_raw.date, format="%m/%d/%Y")),]
sorted_data

#Plotting ratio
ggplot(sorted_data, aes(x=covmeta_raw.date, y=ratio*100)) + geom_point() + labs(x = "Date",y ="Lineage")

#Smoothed Moving Average 7 days
mvg.avg7 = SMA(sorted_data$ratio, n = 7)
plot.ts(mvg.avg7*100, ylab="B.1.1.7 prevelance")
mvg.avg7

sorted_data %>% ggplot(aes(x= as.Date(covmeta_raw.date, format='%m/%d/%Y'), y = mvg.avg7*100)) + geom_line(group=1) + labs(x = " Collection Date",y ="B.1.1.7 prevelance", title= "B.1.1.7 lineage in USA over time") 





```


# Plotting Frequency estimation of B.1.1.7 Lineage in Washington

### Estimation of the ratio of occurences of B.1.1.7 lineage in Washington State

```{r}


dfs = covtry %>% filter(covmeta_raw.division == "Washington")

new_dfs <- dfs %>%
  drop_na() %>% # if we want "ignore" NAs when counting
  group_by(covmeta_raw.date) %>%
  summarise(ratio = sum(covmeta_raw.pango_lineage=="B.1.1.7")/cumsum(n())+0.0001)
new_dfs

new_dfs %>% group_by(covmeta_raw.date)
wsorted_data = new_dfs[order(as.Date(new_df$covmeta_raw.date, format="%m/%d/%Y")),]
wsorted_data 
na.omit(wsorted_data)
#Plotting ratio
ggplot(wsorted_data, aes(x=covmeta_raw.date, y=ratio*100)) + geom_line(group=1) + labs(x = "Date",y ="Lineage")

#Smoothed Moving Average 7 days
wmvg.avg7 = SMA(na.omit(wsorted_data$ratio), n = 7)
plot.ts(wmvg.avg7*100, ylab="B.1.1.7 prevelance")
na.omit(wmvg.avg7)


wsorted_data[1:111,] %>% ggplot(aes(x= as.Date(covmeta_raw.date, format='%m/%d/%Y'), y =wmvg.avg7*100)) + geom_line()+ labs(x = "Date",y ="B.1.1.7 prevelance", title= "B.1.1.7 lineage in Washington over time") + scale_x_date(date_breaks = "1 month", date_labels = "%b %d") + geom_smooth(method ="loess")


```









