---
title: "06-23-21-Rt_dataset_tidy.Rmd"
author: "Eslam Abousamra"
date: "6/23/2021"
output: word_document
---

```{r}
library(tidyverse)
library(openintro)
```

## Prepare case counts

Load CDC case counts for the United States.

```{r}
cdc_raw = read_csv(
  "../cdc-data.csv",
  col_select = c(submission_date, state, new_case)
)
```

Prepare CDC case counts for analysis.
Convert state abbreviation to full names.
Map other non-standard abbreviations to corresponding state/territory names, where possible (e.g., "NYC" maps to "New York").
Group by full state/territory name and sum counts by these names to make sure we have a single value per date and state.

```{r}
cdc_mod = cdc_raw %>% 
  rename(date = submission_date) %>% 
  mutate(newstate = abbr2state(state)) %>%
  mutate(
    newstate = case_when(
      state == "AS" ~ "American Samoa",
      state == "FSM" ~ "Federated States of Micronesia",
      state == "GU" ~ "Guam",
      state == "MP" ~ "Commonwealth of the Northern Mariana Islands",
      state == "NYC" ~ "New York",
      state == "PR" ~ "Puerto Rico",
      state == "PW" ~ "Republic of Palau",
      state == "RMI" ~ "Republic of the Marshall Islands",
      state == "VI" ~ "U.S. Virgin Islands",
      TRUE ~ as.character(newstate)
    )
  ) %>%
  group_by(date, newstate) %>% 
  summarize(new_case = sum(new_case)) %>%
  ungroup() %>%
  rename(division = newstate) %>%
  filter(!is.na(division)) %>%
  mutate(date = as.Date(date, format("%m/%d/%Y"))) 

cdc_mod
```

```{r}
class(cdc_mod$date)
view(cdc_mod)
```

## Prepare genomic metadata

Load GISAID metadata for the United States.
These data include Pango lineage annotations that allow us to calculate the frequencies of variants of concern for each date.

```{r}
init_df = read_tsv(
  "../metadata_gisaid_usa_since_2021-01-01.tsv.gz",
  col_select = c(date, division, pango_lineage)
)
```

Calculate counts and frequencies for specific variants of concern per date and state.

```{r}
var_of_concern = c("B.1.1.7", "B.1.351", "P.1", "B.1.427", "B.1.617.2")

#tidying and prepping the dataset
init_df_new = init_df %>%
  mutate(pango_lineage = case_when(pango_lineage %in% var_of_concern ~ pango_lineage, TRUE ~ "other")) %>%  
  group_by(date, division) %>%
  count(date, pango_lineage, division)%>%
  mutate(totalcount = sum(n, na.rm = FALSE))%>% 
  ungroup() %>% 
  mutate(freq = n/totalcount) %>%
  complete(date, pango_lineage, division,
             fill = list(N = 0, freq = 0)) %>%
  mutate(
    n = replace_na(n, 0),
    freq = replace_na(freq, 0),
    totalcount = replace_na(totalcount, 0)
  )

init_df_new
```

## Merge case counts and genomic data

Merge case counts and genomic data on date and state.
Each row represents a lineage for a specific variant of concern at a given date in a specific state.
These rows redundantly encode case counts for the corresponding date and state.

```{r}
initial_mod = init_df_new %>% left_join(cdc_mod) %>% filter(!is.na(new_case))
view(initial_mod)
```

Save merged data frame for downstream analyses.

```{r}
write_csv(
  initial_mod,
  "counts_and_frequencies_per_date_state_and_lineage.csv"
)
```
